Seeing Double: Transient Dips and Photometric Binarity in PTFO$\,$8-8695

### 2020-04-22 paper/ms.tex
update paper

\subsection{Short period dip}
The TESS lightcurve shows a dip that lasts about 45 minutes, and seems
to re-occur every 10.74 hours
(Figures~\ref{fig:splitsignal},~\ref{fig:splitsignalii},~\ref{fig:phasefold}).
The dip duration is roughly the same as that observed by previous
investigators \citep{van_eyken_ptf_2012,yu_tests_2015}.  The 1.2\%
depth is similar to what has been observed in the near-infrared
\citep{onitsuka_multicolor_2017}.  However the dip depth seems likely
to have evolved over time between being not present at all, to a
maximum of $\approx$5\% \citep[{\it
e.g.},]{koen_multicolour_2015,yu_tests_2015,tanimoto_evidence_2020}.

@@

\subsubsection{Planetary interpretation}
If there were a giant planet transiting \ptfo, it would tidally
distort the host star, and cause ellipsoidal photometric modulations
that also peak at quadrature \citep[see][]{shporer_astrophysics_2017}.
Interpreting the second cosine harmonic as being caused by a planet,
it would imply a minimum planet mass $M_{\rm p} \sin i$ of
$3.8\,M_{\rm Jup}$.  For this estimate, we assumed $R_\star = 1.39
R_\odot$, and $M_\star = 0.39 M_\odot$ \citep{van_eyken_ptf_2012}.
This ellipsoidal amplitude is larger than the typical
modulations induced by close-in giant planets because the host star is
puffy, and still on the pre-main-sequence.

@@

When physical explanations are not immediately clear, taxonomy can be
a useful exercise.  To our knowledge, of order 10 lightcurves with
similar morphologies have been seen in surveys of low-mass weak-lined
T Tauri stars in regions including $\rho$ Oph, Upper Sco, Taurus, and perhaps
the Pleiades
\citep{rebull_rotation_2016,david_transient_2017,stauffer_orbiting_2017,stauffer_rotevol_2018,rebull_usco_2018,rebull_rotation_2020}.
These surveys were performed using K2 \citep{howell_k2_2014}.  We
downloaded a few of these lightcurves from MAST, opting for the
EVEREST reductions \citep{luger_everest_2016,luger_update_2018}.  They
are plotted in Figure~\ref{fig:brethren}

These lightcurves have been classified phenomenologically as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.
This class of objects has the following properties.
\begin{enumerate}
  \item Weak-lined T Tauri stars.
  \item Spectral type typically M2-M5 ({\it e.g.},
    \citet{rebull_usco_2018}~Figure~20).
  \item Ages $\lesssim$ 100$\,$Myr.
  \item Stars typically lack spectroscopic accretion indicators 
  \item Phased lightcurves show shallow, angular dips, usually
    superposed on large-amplitude smooth variability (interpreted as
    stellar rotation).
  \item Rotation is usually rapid ($P<1\,{\rm day}$).
  \item Rarely any infrared excess (never any W4 detection; only a few
    cases with W3 detctions).
  \item Can show multiple dips per cycle ({\it e.g.}, EPIC 203692610,
    RIK-210).
  \item Dip depths, durations, and phasescan vary over timescales of days
    ({\it e.g.}, EPIC 204143627).
  \item Dip depths can change after flares ({\it e.g.}, RIK-210.
  \item Rare ($\lesssim 1\%$ of relevant stars) at a population level
    \citep{rebull_usco_2018}.
\end{enumerate}
The 10.74 hour signal in \ptfo\ meets essentially all of these
criteria.

@@

Given the evidence, we believe that PTFO$\,$8-8695 is a binary M dwarf
in which one star shows the ``long'' rotation signal, and the other is
showing ``transient dipping'' that has also been observed in other
young M dwarfs.

Many other young M-dwarf photometric binaries have been observed in
{\it e.g.,} Upper Sco and the Pleiades to show multiple periods
\citep{rebull_usco_2018,stauffer_rotevol_2018}.  Typically, the
periods are both rotational modulation.

However, occasionally one or both periods can be ``scallop-shell''
variability, {\it e.g.}, EPIC 203956650 in $\rho$Oph
\citep{rebull_usco_2018}.

There are

The gas in the disks is gone (CITE).
The upper limits on the SED from \citet{yu_tests_2015} imply X, Y, Z.
The stars are therefore presumably no longer ``magnetically locked''
to their disks.
This is consistent with the $\approx$half-day periodicities of both
signals.
In the broader context of rapidly rotating young M-dwarfs, ``magnetic
locking'' causes classical disked T-Tauri stars tend to rotate more
{\it slowly}, and almost none have periods less than two days
\citep[{\it e.g.},][]{rebull_rotation_2020}.

The main physical question is what is causing the ``transient
dipping''. This is an unsolved problem not only for \ptfo\ but also
for an entire class of young rapidly rotating M-dwarfs

>>>>

\subsection{Short period dip}

The TESS lightcurve shows a dip that lasts about 45 minutes, and seems
to re-occur every 10.74 hours
(Figures~\ref{fig:splitsignal},~\ref{fig:splitsignalii},~\ref{fig:phasefold}).
The dip duration is roughly the same as that observed by previous
investigators \citep{van_eyken_ptf_2012,yu_tests_2015}.  The 1.2\%
depth is similar to what has been observed in the near-infrared
\citep{onitsuka_multicolor_2017}.  However the dip depth seems likely
to have evolved over time between being not present at all, to a
maximum of $\approx$5\% \citep[{\it
e.g.},]{koen_multicolour_2015,yu_tests_2015,tanimoto_evidence_2020}.

@@

\subsubsection{Planetary interpretation}
If there were a giant planet transiting \ptfo, it would tidally
distort the host star, and cause ellipsoidal photometric modulations
that also peak at quadrature \citep[see][]{shporer_astrophysics_2017}.
Interpreting the second cosine harmonic as planet-induced tidal distortion,
it would imply a minimum planet mass $M_{\rm p} \sin i$ of
$3.8\,M_{\rm Jup}$.  For this estimate, we assumed $R_\star = 1.39
R_\odot$, and $M_\star = 0.39 M_\odot$ \citep{van_eyken_ptf_2012}.
This ellipsoidal amplitude is larger than the typical
modulations induced by close-in giant planets because the host star is
puffy, and still on the pre-main-sequence.

@@

When physical explanations are not immediately clear, taxonomy can be
a useful exercise.  To our knowledge, of order 10-20 lightcurves with
similar morphologies have been seen in surveys of low-mass weak-lined
T Tauri stars in regions including $\rho$ Oph, Upper Sco, Taurus, and perhaps
the Pleiades
\citep{rebull_rotation_2016,david_transient_2017,stauffer_orbiting_2017,stauffer_rotevol_2018,rebull_usco_2018,rebull_rotation_2020}.
These surveys were performed using K2 \citep{howell_k2_2014}.  We
downloaded a few of these lightcurves from MAST, opting for the
EVEREST reductions \citep{luger_everest_2016,luger_update_2018}.  They
are plotted in Figure~\ref{fig:brethren}.

These lightcurves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.
As \citet{stauffer_orbiting_2017} explain,
these  objects are morphologically distinct from ``scallop shell' lightcurves,
and are present in stars at more advanced evolutionary disk stages than
the ``dipper'' stars \citep{ansdell_young_2016,cody_manyfaceted_2018}.
The persistent and transient flux dip stars have the following properties.
\begin{enumerate}
  \item Weak-lined T Tauri stars.
  \item Spectral type typically M2-M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item Ages $\lesssim$ 100$\,$Myr.
  \item Stars typically lack spectroscopic accretion indicators 
  \item Phased lightcurves show shallow, angular dips, usually
    superposed on large-amplitude smooth variability. The latter
    is interpreted as stellar rotation.
  \item Rotation is usually between 0.5 to 2.0 days.
  \item Rarely any infrared excess (never any W4 detection; only a few
    cases with W3 detctions).
  \item Can show multiple dips per cycle ({\it e.g.}, EPIC 203692610,
    RIK-210).
  \item Dip depths, durations, and phases can vary over just a few cycles
    ({\it e.g.}, EPIC 204143627).
  \item Dip depths can change after flares.
  \item Rare at a population level: $\lesssim 1\%$ of relevant stars
    \citep{rebull_usco_2018}.
\end{enumerate}
The 10.74 hour signal in \ptfo\ meets essentially all of these
criteria.

A crucial additional point concerning the transient flux dips is
that their dip durations seem to scale linearly with
their periods \citep[][Figure~26]{stauffer_orbiting_2017}.
In the usual idealized limits, the transit duration of a point source
across the stellar disk scales as $T_0 \propto R_\star (P/M_\star)^{1/3}$
\citep{winn_exoplanet_2010}.
While the $\approx$0.5-day transient flux dip stars have dip
durations consistent with point sources, at longer periods of
1 to 5 days, the dip durations become many hours, and far
too long for a point-source origin.

@@

The evidence for binarity in \ptfo\ is as follows.  First, the star is
photometrically twice as bright as stars of the same color in its
kinematic group (Figure~\ref{fig:gaia}.  Second, it shows two distinct
photometric signals.  These two points alone are typically sufficient
to imply binarity \citep{stauffer_rotevol_2018}.  For the case of
\ptfo, there is a third line of evidence: the Gaia astrometry shows a
poor goodness-of-fit to a single-source model.  While this could be
caused by an additional companion or stellar variability, cluster
members that are equally if not more variable still show \ptfo\ to be
an outlier.  Therefore the astrometric excess is a suggestive third
line of evidence for binarity in \ptfo.  We take these three points
combined to imply that \ptfo\ is probably a binary.

% rumination: is the lack of an infrared excess even significant? In
% USco, we know late-M dwarfs show them. What about Orion? Distance is
% 4x, at least...
% Hernandez et al 2007a and 2007b show that there are some infrared
% excesses seen with Spitzer. But their completeness for fainter M
% dwarfs may be worse -- they see a disk fraction peaking at M0,
% instead of later as seen by e.g., Barenfeld+2016.
Based on the lack of an infrared excess seen by \citet{yu_tests_2015},
the primordial gas disks seem to be have been depleted around both
stars, and so the stars are no longer magnetically locked to their
disks.  This is consistent with the $\approx$half-day periodicities of
both signals: young disked M dwarfs typically rotate with periods of
two days or more due to magnetic locking \citep[{\it
e.g.},][]{rebull_rotation_2020}.  If the two stars are within
$\approx$50$\,$AU of each other, as required by the NIRC2 adaptive
optics imaging, then it would also be expected that the stars would
truncate the outer edges of their respective disks, in a manner seen
at the population level in exoplanetary systems
\citep{kraus_impact_2016,moe_impact_2019}.

The main physical question remaining is clearly what causes the
transient dipping. We have argued that this is an unsolved problem not
only for \ptfo\ but also for an emerging class of similar young
rapidly rotating M-dwarfs.

The explanation offered by \citet{stauffer_orbiting_2017} for this
variability class is that warm clouds of gas and dust orbit the star
at the Keplerian corotation radius.

Why???

What are the forces at play?

FIXME

Given the evidence for binarity, we believe that PTFO$\,$8-8695 is a
binary M dwarf in which one star shows the 11.96~hour rotation signal,
and the other shows a combination of transient dipping and stellar
rotation, similar to that observed in other young M dwarfs.  The
combined transient dipping and stellar rotation at the 10.74~hour
cycle was only detected with the TESS data.
